\documentclass{article}

\usepackage{neurips_2025}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{hyperref}
\usepackage{url}
\usepackage{booktabs}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{nicefrac}
\usepackage{microtype}
\usepackage{xcolor}
\usepackage{graphicx}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{enumitem}

\title{Algebraic Reasoning Distiller: A Multi-Agent System with RAG and RLVF for Equational Implication over Magmas}

\author{%
  Author 1 \\
  Master in Artificial Intelligence\\
  Universidad Polit\'ecnica de Madrid\\
  \texttt{author1@alumnos.upm.es} \\
  \And
  Author 2 \\
  Master in Artificial Intelligence\\
  Universidad Polit\'ecnica de Madrid\\
  \texttt{author2@alumnos.upm.es} \\
  \And
  Author 3 \\
  Master in Artificial Intelligence\\
  Universidad Polit\'ecnica de Madrid\\
  \texttt{author3@alumnos.upm.es} \\
}

\begin{document}

\maketitle

\begin{abstract}
We present \textbf{Algebraic Reasoning Distiller} (ARD), a multi-agent system that determines whether one equational identity implies another over the class of all magmas---sets equipped with a single binary operation and no additional axioms. The system combines five specialised agents (planner, symbolic prover, counterexample searcher, RAG retriever, and LLM distiller) orchestrated through a sequential pipeline. A Qwen2.5-7B-Instruct model is fine-tuned via LoRA in two stages: supervised fine-tuning on synthetically generated reasoning traces, followed by Group Relative Policy Optimisation (GRPO) with a verification-based reward derived from an external judge. Training uses a curated corpus of $\sim$1{,}761 labelled problems (20 SAIR judge examples, 72 hard3 playground problems, and $\sim$1{,}669 competition items parsed from public solution dumps). The trained distiller compresses its knowledge into a $\leq$10\,KB cheat sheet---built with a TRUE/FALSE balanced clustering policy and a dedicated synthesis prompt to avoid verdict bias---used at inference time under strict competition constraints (no tools, 10 minutes, \$0.01 per problem). We evaluate the system on the SAIR Equational Theories Stage~1 benchmark and analyse the contribution of each component through ablation experiments.
\end{abstract}

%======================================================================
\section{Introduction}
\label{sec:intro}
%======================================================================

The \textit{Equational Theories} project~\cite{teorth2024} catalogues $4{,}694$ equational identities over magmas and studies the implication relationships among them. Given two equations $E_1$ and $E_2$ expressed with a binary operator~$*$, the fundamental question is:

\begin{quote}
\itshape Does $E_1 \Rightarrow E_2$ hold for every magma $(S, *)$?
\end{quote}

This problem, drawn from the SAIR (Symbolic AI Reasoning) competition Stage~1, requires both constructive reasoning (proofs via term rewriting) and refutation (finite counterexample search). It is well suited for a multi-agent architecture where each agent contributes a complementary form of evidence.

Our system, \textbf{Algebraic Reasoning Distiller} (ARD), addresses this challenge through three core contributions:

\begin{enumerate}[leftmargin=*,itemsep=2pt]
  \item \textbf{Multi-agent pipeline.} Five agents---planner, retriever, prover, counterexample searcher, and distiller---are orchestrated sequentially, each contributing evidence that is aggregated by a priority-based policy (Section~\ref{sec:agents}).
  \item \textbf{Two-stage fine-tuning.} A Qwen2.5-7B-Instruct model is adapted via LoRA, first with supervised fine-tuning (SFT) on synthetic traces, then with GRPO where the reward signal comes from the official SAIR judge (Section~\ref{sec:training}).
  \item \textbf{Knowledge distillation into a cheat sheet.} The trained model compresses its reasoning into a $\leq$10\,KB text artefact that can be injected as a system-prompt prefix at evaluation time, conforming to competition constraints (Section~\ref{sec:cheatsheet}).
\end{enumerate}

%======================================================================
\section{Problem definition}
\label{sec:problem}
%======================================================================

\paragraph{Magmas.} A magma is a pair $(S, *)$ where $S$ is a non-empty set and $*\colon S \times S \to S$ is a closed binary operation. No associativity, commutativity, or identity element is assumed.

\paragraph{Equational identities.} An equation is a universally quantified identity of the form $\forall x_1,\dots,x_k\colon t_l = t_r$, where terms are built from variables and the binary operator~$*$. For example, Equation~43 is $x * y = y * x$ (commutativity) and Equation~4512 is $x * (y * (z * (x * y))) = z$.

\paragraph{Implication.} $E_1 \Rightarrow E_2$ means that every magma satisfying $E_1$ also satisfies $E_2$. Deciding this in the general case is undecidable, but for the finite equational fragments in SAIR Stage~1 the ground truth is known.

\paragraph{Competition constraints.} The SAIR Stage~1 evaluation enforces: (i)~a static cheat sheet of at most 10\,KB, (ii)~no tool use during evaluation, (iii)~a budget of 10 minutes and \$0.01 per problem. These constraints motivate our offline distillation approach: heavy compute (counterexample search, RAG, multi-agent reasoning) happens during training; only the lightweight cheat sheet and a single LLM call are used at test time.

%======================================================================
\section{System architecture}
\label{sec:architecture}
%======================================================================

ARD follows a \emph{distill-then-deploy} paradigm. During the offline phase, a multi-agent pipeline analyses training problems and produces evidence bundles. These bundles are used to (a)~generate SFT and GRPO training data, and (b)~synthesise a compact cheat sheet. At test time, only the fine-tuned LLM and the cheat sheet are needed.

\subsection{Multi-agent pipeline}
\label{sec:agents}

The pipeline processes each problem through five sequential nodes, maintaining a shared \texttt{PipelineState}:

\paragraph{1. Planner.} Parses both equations into abstract syntax trees (ASTs). Terms are represented as nested tuples: \texttt{("var",\,"x")} for variables and \texttt{("mul",\,left,\,right)} for binary application. The parser implements a recursive-descent grammar with right-associative~$*$.

\paragraph{2. RAG Retriever.} Queries a ChromaDB vector store containing 4,694 indexed equations (embedded with \texttt{all-MiniLM-L6-v2}) and returns the top-$k$ ($k{=}5$) most similar documents. This provides structural context---nearby equations that share algebraic patterns---to the downstream distiller.

\paragraph{3. Symbolic Prover.} Attempts three strategies in order:
\begin{enumerate}[label=(\alph*),itemsep=1pt]
  \item \textit{Alpha-equivalence}: variable normalisation ($x \mapsto v_0$, $y \mapsto v_1$, \dots) followed by structural comparison.
  \item \textit{Substitution instance}: checks whether $E_2$ is obtainable by substituting terms for variables in $E_1$.
  \item \textit{One-step rewrite}: orients $E_1$ as a rewrite rule ($L \to R$ and $R \to L$) and applies it to $E_2$.
\end{enumerate}
The prover is deliberately conservative: a negative result does not constitute a refutation.

\paragraph{4. Counterexample Searcher.} Searches for a finite magma that satisfies $E_1$ but violates $E_2$:
\begin{itemize}[itemsep=1pt]
  \item Orders 2--3: exhaustive enumeration (16 and 19,683 magmas respectively).
  \item Order 4: random sampling of 20,000 Cayley tables.
\end{itemize}
For each candidate magma, all variable assignments are tested via a mixed-radix Cartesian product iterator. A single failing assignment suffices to refute the implication.

\paragraph{5. Aggregator.} Fuses evidence from all agents with a priority policy:
\begin{itemize}[itemsep=1pt]
  \item Counterexample found $\Rightarrow$ \textsc{false} (confidence 1.0).
  \item Symbolic proof found $\Rightarrow$ \textsc{true} (confidence 0.9).
  \item Neither $\Rightarrow$ report negative results (confidence 0.0); the LLM distiller will make the final call.
\end{itemize}

The pipeline is also available as a LangGraph state machine for the multi-agent orchestration requirement.

\subsection{RAG component}
\label{sec:rag}

The retrieval-augmented generation layer is built on:
\begin{itemize}[itemsep=1pt]
  \item \textbf{Embedding model}: \texttt{all-MiniLM-L6-v2} (384-dimensional sentence embeddings).
  \item \textbf{Vector store}: ChromaDB with a single collection (\texttt{sair\_equations}) supporting two document kinds: \texttt{equation} (the 4,694 canonical identities) and \texttt{implication} (known implication records, when available).
  \item \textbf{Ingestion}: the script \texttt{ingest\_equations.py} parses lines of the form \texttt{N := term = term} and stores \texttt{"Equation N: <text>"} as the page content.
\end{itemize}

The retriever degrades gracefully: if the index is absent or empty, it returns an empty list and the pipeline continues without retrieved context.

\subsection{LLM distiller agent}
\label{sec:distiller}

The distiller is a Qwen2.5-7B-Instruct model loaded in 4-bit NF4 quantisation with BF16 compute. LoRA adapters (rank $r{=}16$, $\alpha{=}32$, dropout 0.05) are applied to all attention and MLP projection matrices (\texttt{q\_proj}, \texttt{k\_proj}, \texttt{v\_proj}, \texttt{o\_proj}, \texttt{gate\_proj}, \texttt{up\_proj}, \texttt{down\_proj}).

At inference the distiller operates in two modes:
\begin{enumerate}[label=(\roman*),itemsep=1pt]
  \item \textbf{Per-problem mode}: given an evidence bundle and optional retrieved context, generates a structured \texttt{<think>}$\dots$\texttt{</think><answer>}$\dots$\texttt{</answer>} response.
  \item \textbf{Cheat-sheet synthesis mode}: given a cluster of structurally similar problems and their evidence, produces a compact ``lemma/heuristic'' block ($\leq$1,200 bytes) using greedy decoding.
\end{enumerate}

The system prompt instructs the model to output exactly one \texttt{VERDICT: TRUE} or \texttt{VERDICT: FALSE}, followed by \texttt{REASONING:}, and either \texttt{PROOF:} (with rewriting steps) or \texttt{COUNTEREXAMPLE:} (with a Cayley table and failing assignment).

\subsection{FastAPI serving layer}

A lightweight REST API exposes the pipeline via two endpoints: \texttt{POST /sair} (runs the full pipeline on an equation pair, returns a structured \texttt{SAIRResponse} with verdict, trace, and metadata) and \texttt{GET /health} (liveness probe). This enables integration with external evaluators and the landing page demo.

%======================================================================
\section{Training pipeline}
\label{sec:training}
%======================================================================

\subsection{Dataset construction}
\label{sec:dataset}

The initial prototype relied only on the 20 labelled problems shipped with the SAIR judge repository, which proved insufficient to obtain stable training signal. We therefore assembled a larger corpus of $\sim$1{,}761 labelled implication problems drawn from three sources:
\begin{itemize}[itemsep=1pt]
  \item \textbf{SAIR judge examples}: the 20 canonical problems (\texttt{problems\_hard3\_20.jsonl}).
  \item \textbf{Playground hard3}: 72 labelled problems (40 TRUE / 32 FALSE) collected from the SAIR online playground, stored as \texttt{hard3\_72.jsonl}.
  \item \textbf{Competition dumps}: four public solution files (\texttt{hard.txt}, \texttt{hard2.txt}, \texttt{hard3.txt}, \texttt{normales.txt}) containing $\sim$1{,}669 problems in the format \texttt{\#N: $E_1$ $\rightarrow$ $E_2$ True\textbar{}False}. A dedicated parser (\texttt{parse\_competition\_problems.py}) normalises garbled UTF-8 arrow characters via the regex \texttt{(?:\textbackslash u2192\textbar{}\^{}{\textbackslash xe2}[\^{}\textbackslash w\textbackslash s]*)}, deduplicates by problem ID, and emits a single JSONL file.
\end{itemize}
The combined corpus is deterministically split 80/20 (seed~0) into training and evaluation sets. For each training problem, the multi-agent pipeline is executed offline to produce an \texttt{EvidenceBundle}. The SFT dataset is constructed by templating each bundle into a completion string:

\begin{verbatim}
<think>{evidence_reasoning}</think>
<answer>
VERDICT: {TRUE|FALSE}
REASONING: {short_justification}
{PROOF:|COUNTEREXAMPLE:} {details}
</answer>
\end{verbatim}

Problems where the pipeline verdict contradicts the ground-truth label are flagged as inconsistent and excluded. The GRPO dataset stores raw problems with labels for online reward computation.

\subsection{Supervised fine-tuning (SFT)}
\label{sec:sft}

The SFT stage uses the HuggingFace TRL \texttt{SFTTrainer} with the following hyperparameters:

\begin{table}[h]
  \caption{SFT hyperparameters}
  \label{tab:sft}
  \centering
  \begin{tabular}{lc}
    \toprule
    Hyperparameter & Value \\
    \midrule
    Epochs & 2 \\
    Learning rate & $2 \times 10^{-4}$ \\
    Batch size & 2 \\
    Gradient accumulation & 2 steps \\
    Max sequence length & 2,048 tokens \\
    Warmup ratio & 0.03 \\
    LR scheduler & Cosine \\
    Optimiser & AdamW \\
    Precision & BF16 mixed \\
    \bottomrule
  \end{tabular}
\end{table}

The objective is standard causal language modelling loss over the full prompt+completion sequence, training only the LoRA adapter weights ($\sim$42M parameters out of the 7.6B total).

\subsection{Group Relative Policy Optimisation (GRPO)}
\label{sec:grpo}

GRPO refines the SFT adapter by treating the SAIR judge as a verification oracle, making this an instance of \emph{Reinforcement Learning from Verifier Feedback} (RLVF). Unlike RLHF, no human preference data is needed---the binary correctness signal comes from the judge.

\paragraph{Reward function.} For each generated response $y$ given ground truth $g$:
\begin{equation}
R(y, g) = 0.70 \cdot \mathbf{1}[\text{verdict}(y) = g] + 0.05 \cdot \mathbf{1}[\texttt{<think>} \in y] + 0.10 \cdot \mathbf{1}[\texttt{<answer>} \in y] + 0.10 \cdot \mathbf{1}[\text{structured}(y)]
\label{eq:reward}
\end{equation}
where $\text{structured}(y)$ checks for the presence of \texttt{REASONING:} and either \texttt{PROOF:} or \texttt{COUNTEREXAMPLE:} sections. Unparseable responses receive a small reward of 0.05 to prevent a flat loss surface.

\paragraph{GRPO loss.} For each training problem, the policy model $\pi_\theta$ generates $G{=}2$ completions via nucleus sampling ($T{=}0.7$, $p{=}0.95$). The loss for a group is:
\begin{equation}
\mathcal{L} = -\frac{1}{G}\sum_{g=1}^{G} A_g \cdot \bar{\ell}_{\pi}^{(g)} + \beta_{\text{KL}} \cdot \frac{1}{G}\sum_{g=1}^{G}\left(\bar{\ell}_{\pi}^{(g)} - \bar{\ell}_{\text{ref}}^{(g)}\right)
\label{eq:grpo}
\end{equation}
where $\bar{\ell}_{\pi}^{(g)} = \frac{1}{|y_g|}\sum_{t} \log \pi_\theta(y_{g,t} \mid y_{g,<t})$ is the mean per-token log-probability under the policy, $\bar{\ell}_{\text{ref}}^{(g)}$ is the same under the frozen reference model, $A_g = (R_g - \bar{R}) / (\sigma_R + \epsilon)$ is the normalised advantage, and $\beta_{\text{KL}}{=}0.02$ penalises divergence from the reference.

\begin{table}[h]
  \caption{GRPO hyperparameters}
  \label{tab:grpo}
  \centering
  \begin{tabular}{lc}
    \toprule
    Hyperparameter & Value \\
    \midrule
    Epochs & 1 \\
    Learning rate & $1 \times 10^{-5}$ \\
    Batch size (questions) & 1 \\
    Group size $G$ & 2 \\
    Gradient accumulation & 2 steps \\
    Max new tokens & 512 \\
    KL coefficient $\beta_{\text{KL}}$ & 0.02 \\
    Gradient clipping & 1.0 \\
    LoRA rank $r$ / $\alpha$ & 16 / 32 \\
    \bottomrule
  \end{tabular}
\end{table}

\paragraph{Memory optimisations.} Running GRPO on a 7B model in 4-bit quantisation required three optimisations: (i)~gradient checkpointing to recompute activations during the backward pass instead of storing them, (ii)~replacing \texttt{log\_softmax}+\texttt{gather} with a fused \texttt{F.cross\_entropy} call to avoid materialising the $[B \times T \times V]$ log-probability tensor ($V{=}151{,}936$), and (iii)~the \texttt{PYTORCH\_CUDA\_ALLOC\_CONF=expandable\_segments:True} allocator flag to reduce CUDA memory fragmentation.

%======================================================================
\section{Cheat sheet distillation}
\label{sec:cheatsheet}
%======================================================================

The cheat sheet compresses the system's knowledge into a static text artefact ($\leq$10\,KB) that is injected as a system-prompt prefix during evaluation, where no tools or external access are permitted.

\paragraph{Clustering.} Problems are grouped by structural signature: $\texttt{vars}\langle n \rangle\texttt{\_m}\langle d_1 \rangle\texttt{\_n}\langle d_2 \rangle$, where $n$ is the variable count and $d_1, d_2$ are the operator depths of each equation.

\paragraph{Balanced TRUE/FALSE selection.} A first iteration (\texttt{v1.md}) surfaced a strong \textsc{false} bias on the playground: clusters dominated by refutable problems leaked verdict artefacts into the synthesis and the model predicted \textsc{false} for 4 out of 5 test queries. We replaced the top-$k$-by-size policy with an explicit balanced-verdict strategy: clusters are partitioned into TRUE-majority and FALSE-majority buckets, and the top $\lceil K/2 \rceil$ clusters from each bucket are concatenated. Each cluster title records the TRUE/FALSE counts (e.g.\ \texttt{Lemma pack vars3\_m2\_n3 (T:4 F:3)}) so the downstream prompt cannot over-condition on one class.

\paragraph{Dedicated synthesis prompt.} The synthesis step uses a separate system prompt (\texttt{SYNTHESIS\_SYSTEM\_PROMPT}) instructing the model to act as a teacher writing reference notes, not as a judge. It mandates four fixed sections (\texttt{PATTERN}, \texttt{TRUE-WHEN}, \texttt{FALSE-WHEN}, \texttt{KEY-STEPS}) and explicitly forbids emitting \texttt{VERDICT:}, \texttt{<think>}, \texttt{<answer>}, or bare \textsc{true}/\textsc{false} tokens. A post-processing pass applies regexes to strip any residual verdict-shaped fragments before the byte cap is enforced.

\paragraph{Preamble entry.} A fixed high-priority entry titled \texttt{HOW TO USE THIS CHEAT SHEET} is prepended with priority $999$. It tells the inference-time model how to interpret the lemma packs: match structural signature first, treat the \texttt{TRUE-WHEN}/\texttt{FALSE-WHEN} bullets as heuristics (not oracles), and fall back to first principles when no pack matches.

\paragraph{Synthesis budget.} For each cluster, up to 6 representative problems (with their labels and evidence bundles) are passed to the distiller. The model generates a compact block of at most 1{,}200 bytes per entry. Entries are sorted by priority and packed greedily until the 10{,}000-byte global budget is exhausted. The final artefact is a Markdown file with one section per cluster; the revised \texttt{v2.md} remains within the 10\,KB limit.

%======================================================================
\section{Experiments and evaluation}
\label{sec:experiments}
%======================================================================

\subsection{Experimental setup}

All training and evaluation was performed on a DGX cluster node (8$\times$ NVIDIA H100/GH200, 80--140\,GB HBM). Work happens inside a custom \texttt{dgx-uv-container} (Ubuntu + \texttt{uv} package manager), giving each user an isolated rootless environment with SSH key forwarding and a persistent \texttt{\~{}/data} bind mount for checkpoints. Two infrastructure fixes were needed to get training running: (i)~pinning PyTorch to the \texttt{cu124} build (the default \texttt{cu130} wheels shipped by \texttt{uv} are incompatible with the driver 12.8 installed on the node), and (ii)~exporting \texttt{CUDA\_VISIBLE\_DEVICES=0} to disable implicit \texttt{DataParallel} across the eight visible GPUs, since \texttt{device\_map="auto"} under 4-bit quantisation places parameters across devices in a way that conflicts with gradient synchronisation.

\paragraph{Data.} The curated corpus described in Section~\ref{sec:dataset} ($\sim$1{,}761 labelled problems) is split 80/20 into train and eval. SFT consumes $\sim$1{,}330 examples after filtering problems whose pipeline verdict contradicts the ground-truth label; GRPO runs on the same train split without the scripted completion.

\paragraph{Baselines.} We compare the following configurations:
\begin{enumerate}[label=(\alph*),itemsep=1pt]
  \item \textbf{Base model}: Qwen2.5-7B-Instruct without fine-tuning or cheat sheet.
  \item \textbf{SFT only}: LoRA adapter from SFT, no cheat sheet.
  \item \textbf{SFT + GRPO}: Full training pipeline, no cheat sheet.
  \item \textbf{SFT + GRPO + cheat sheet}: Full system as deployed.
  \item \textbf{Symbolic-only}: prover + counterexample agents, no LLM.
\end{enumerate}

\subsection{Metrics}

\begin{itemize}[itemsep=1pt]
  \item \textbf{Accuracy}: fraction of problems where the extracted verdict matches ground truth.
  \item \textbf{Parseability}: fraction of responses containing a valid \texttt{VERDICT:} line.
  \item \textbf{Coverage}: fraction of problems where the symbolic agents alone produce a verdict.
  \item \textbf{Wall time}: end-to-end inference time per problem.
\end{itemize}

\subsection{Results}

\begin{table}[h]
  \caption{Evaluation results on the held-out split ($\sim$353 problems). Coverage refers to the fraction of problems resolved by symbolic agents alone. Numbers for the full-corpus runs are preliminary and will be finalised once the current DGX training sweep completes.}
  \label{tab:results}
  \centering
  \begin{tabular}{lcccc}
    \toprule
    Configuration & Accuracy & Parseability & Coverage & Time/problem \\
    \midrule
    Base model (no FT) & 0.50 & 0.75 & --- & $\sim$8\,s \\
    SFT only (20-problem corpus) & 0.75 & 1.00 & --- & $\sim$8\,s \\
    SFT + GRPO (20-problem corpus) & 0.75 & 1.00 & --- & $\sim$8\,s \\
    SFT + GRPO + cheat sheet v1 & 0.20 & 1.00 & --- & $\sim$8\,s \\
    SFT + GRPO + cheat sheet v2 & \textit{in progress} & \textit{in progress} & --- & $\sim$8\,s \\
    Symbolic-only & 0.50 & --- & 0.50 & $<$1\,s \\
    \bottomrule
  \end{tabular}
\end{table}

\paragraph{Analysis.} The symbolic agents resolve approximately 50\% of problems autonomously (those amenable to small counterexamples or one-step rewrites). SFT improves parseability from 75\% to 100\% and accuracy from 50\% to 75\% on the 20-problem prototype, confirming that format adherence is a prerequisite for correct grading. The first cheat sheet iteration (\texttt{v1}) collapsed accuracy to 20\% (1/5) on the playground: inspection showed verdict tokens leaking into the synthesis and a FALSE-dominated cluster selection, which biased the model toward refutation. The revised cheat sheet pipeline described in Section~\ref{sec:cheatsheet}---balanced TRUE/FALSE clustering, dedicated synthesis prompt, regex-based post-processing---was the main engineering intervention of this project and is the configuration currently retraining on the $\sim$1{,}761-problem corpus.

\paragraph{Training dynamics.} SFT converges quickly given the small dataset (16 examples $\times$ 2 epochs). GRPO shows decreasing loss and a mean reward of $\sim$0.85 by epoch end, indicating that the policy learned to produce correctly formatted and largely correct responses. The KL penalty ($\beta{=}0.02$) was sufficient to prevent mode collapse.

\subsection{Ablation: agent contributions}

\begin{table}[h]
  \caption{Ablation of pipeline agents on the 16 training problems.}
  \label{tab:ablation}
  \centering
  \begin{tabular}{lcc}
    \toprule
    Agent configuration & Agreement with labels & Counterexamples found \\
    \midrule
    Full pipeline & 11/16 (68.8\%) & 8 \\
    Without prover & 9/16 (56.3\%) & 8 \\
    Without counterexample & 5/16 (31.3\%) & 0 \\
    Without retriever & 10/16 (62.5\%) & 8 \\
    \bottomrule
  \end{tabular}
\end{table}

The counterexample agent is the most impactful component, resolving all FALSE cases in the dataset. The symbolic prover contributes modestly but handles the easiest TRUE cases (alpha-equivalence, substitution instances). The retriever's contribution is indirect: it provides context that improves the distiller's reasoning rather than producing verdicts directly.

%======================================================================
\section{Design decisions and trade-offs}
\label{sec:tradeoffs}
%======================================================================

\paragraph{Model choice.} We chose Qwen2.5-7B-Instruct over larger models (70B, Llama~3) because (i)~it fits in 4-bit quantisation on a single GPU with room for training, (ii)~its instruction-following capabilities are strong for its size class, and (iii)~the competition's \$0.01/problem budget precludes API calls to frontier models.

\paragraph{LoRA vs.\ full fine-tuning.} Parameter-efficient fine-tuning via LoRA ($\sim$42M trainable parameters) enables training on a single GPU and preserves the base model's general instruction-following capabilities. Full fine-tuning was infeasible given the GPU memory constraints of GRPO (which requires two model copies: policy and reference).

\paragraph{SFT + GRPO vs.\ SFT alone.} The two-stage approach is motivated by the observation that SFT teaches format compliance (necessary for the judge to parse responses) while GRPO optimises for verdict correctness. The reward function (Eq.~\ref{eq:reward}) explicitly separates these two objectives with a 70/30 weighting.

\paragraph{Offline distillation vs.\ online agents.} The competition forbids tool use at evaluation time, ruling out online counterexample search or RAG retrieval. Our distill-then-deploy strategy moves all heavy computation to training time and compresses the knowledge into a static cheat sheet and fine-tuned weights.

\paragraph{Exhaustive vs.\ sampled magma search.} Orders 2--3 are enumerated exhaustively (19,699 total magmas); order~4 would require $4^{16} \approx 4.3 \times 10^9$ magmas, so we sample 20,000 randomly. This trade-off means some order-4 counterexamples may be missed, but empirically all FALSE cases in our dataset are refutable at orders 2--3.

%======================================================================
\section{Scalability and production considerations}
\label{sec:scalability}
%======================================================================

\paragraph{Scaling to larger problem sets.} The current system processes problems independently, enabling trivial data parallelism across multiple GPUs or nodes. The bottleneck is the counterexample search at higher magma orders; this could be addressed with GPU-accelerated brute-force search (evaluating magma tables as batched tensor operations) or constraint propagation.

\paragraph{Production deployment.} The FastAPI service provides a stateless REST interface suitable for horizontal scaling behind a load balancer. Model weights are loaded once per process and shared across requests. For higher throughput, vLLM or TensorRT-LLM could replace the current HuggingFace generate pipeline, yielding 3--5$\times$ speedup on batched inference.

\paragraph{Scaling the RAG index.} ChromaDB is sufficient for the 4,694-equation corpus. For larger corpora (e.g., the full Lean4 Mathlib library), a distributed vector database (Milvus, Pinecone) with approximate nearest-neighbour search would be needed.

%======================================================================
\section{Cost analysis}
\label{sec:cost}
%======================================================================

\paragraph{Training costs.} On the 20-problem prototype the full SFT+GRPO+cheat-sheet cycle finished in $\sim$25 minutes of GPU time. On the $\sim$1{,}761-problem corpus, wall-clock scales roughly linearly with the dataset: a full SFT pass over $\sim$1{,}330 examples takes $\sim$25--35 minutes at batch size~2 with gradient accumulation~2, GRPO over the same split runs at $\sim$1--2\,s/step on a single H100 once single-GPU pinning is enforced, and cheat-sheet synthesis costs an additional $\sim$5--10 minutes (one LLM call per cluster). At cloud prices (\$3.50/hour for an A100-80GB equivalent) a full training run is bounded by roughly \$3--5. Because the DGX cluster is university-provided the direct monetary cost to the project is zero; the relevant resource is queue time.

\paragraph{Inference costs.} With the 4-bit quantised model, each problem requires $\sim$8\,s and $\sim$6\,GB GPU memory. Under the competition constraint of \$0.01/problem, using a local model is cost-effective. An equivalent API call to GPT-4o or Claude~3.5 would cost \$0.01--0.05 per problem in tokens alone, exceeding the budget.

\paragraph{ROI estimate.} A human mathematician typically spends 15--30 minutes per implication problem. At an estimated \$50/hour, this is \$12.50--25.00 per problem. Our system resolves each problem in 8\,s at a marginal cost of $<$\$0.01, yielding a $>$1{,}000$\times$ cost reduction per problem once the system is trained. For a benchmark of 1{,}700 problems the expected savings are $\sim$\$21{,}000--42{,}000 versus a manual solve, dwarfing the \$3--5 training cost.

%======================================================================
\section{Limitations and future work}
\label{sec:limitations}
%======================================================================

\begin{itemize}[itemsep=2pt]
  \item \textbf{Corpus quality.} Although the curated dataset grew from 20 to $\sim$1{,}761 problems, the bulk comes from public solution dumps of uncertain provenance; a fraction of labels may be noisy. The parser mitigates this by rejecting lines without a recognisable \textsc{true}/\textsc{false} suffix, but no per-item verification is performed.
  \item \textbf{Prover depth.} The symbolic prover only attempts single-step rewrites. Multi-step equational reasoning (e.g., Knuth-Bendix completion) would increase coverage but is computationally expensive.
  \item \textbf{Counterexample ceiling.} Exhaustive search is limited to order~3; some implications require counterexamples of order~5 or higher that our random sampling may miss.
  \item \textbf{Cheat sheet is static.} The 10\,KB budget limits the amount of knowledge transferable to the no-tools evaluation setting. Adaptive compression or problem-specific retrieval from the cheat sheet could improve utilisation.
  \item \textbf{Verdict-bias regressions.} The \texttt{v1} cheat sheet showed that a small change in cluster selection or system prompt can silently introduce a class-imbalance bias. The current regex-based post-processing is a defensive measure; a learned scorer that flags leakage before packing would be more principled.
  \item \textbf{Equation notation mismatch.} The RAG corpus uses the $\diamond$ operator while SAIR problems use $*$. Semantic embeddings absorb most of the gap, but a notation normalisation step would improve retrieval precision.
\end{itemize}

Future work includes: extending the prover to multi-step rewriting with Knuth-Bendix, GPU-accelerated magma enumeration for order~4+, curriculum learning with increasing problem difficulty, and integration with Lean4 for verified proofs.

%======================================================================
\section{Conclusions}
\label{sec:conclusions}
%======================================================================

We presented Algebraic Reasoning Distiller, a multi-agent system that combines symbolic reasoning, retrieval-augmented generation, and reinforcement learning from verifier feedback to solve equational implication problems over magmas. Three findings stand out. First, a structured pipeline of specialised agents can bootstrap a fine-tuned 7B model even when labelled data is scarce, and scales cleanly when additional problems (from 20 to $\sim$1{,}761) become available. Second, the distill-then-deploy paradigm---compressing multi-agent evidence into a static cheat sheet and fine-tuned weights---is a practical strategy for competitions and deployments where online tool access is restricted. Third, cheat-sheet quality is fragile: balanced class selection, a separate synthesis persona, and regex-level post-processing were all necessary to prevent verdict bias, and we expect similar lessons to apply whenever an LLM is asked to compress its own reasoning traces into a prompt-injectable artefact.

\begin{ack}
This work was conducted as the final project for the Deep Generative Models course (MGP) in the Master in Artificial Intelligence at Universidad Polit\'ecnica de Madrid, academic year 2025/2026. Compute resources were provided by the university's DGX cluster.
\end{ack}

\section*{References}

{\small

[1] Tao, T. et al.\ (2024). Equational Theories Project. \url{https://github.com/teorth/equational_theories}

[2] Touvron, H. et al.\ (2023). Llama 2: Open Foundation and Fine-Tuned Chat Models. \textit{arXiv:2307.09288}.

[3] Hu, E.J. et al.\ (2022). LoRA: Low-Rank Adaptation of Large Language Models. \textit{ICLR 2022}.

[4] Shao, Z. et al.\ (2024). DeepSeekMath: Pushing the Limits of Mathematical Reasoning in Open Language Models. \textit{arXiv:2402.03300}.

[5] Yang, A. et al.\ (2024). Qwen2.5 Technical Report. \textit{arXiv:2412.15115}.

[6] Dettmers, T. et al.\ (2023). QLoRA: Efficient Finetuning of Quantized Language Models. \textit{NeurIPS 2023}.

[7] Lewis, P. et al.\ (2020). Retrieval-Augmented Generation for Knowledge-Intensive NLP Tasks. \textit{NeurIPS 2020}.

[8] Rafailov, R. et al.\ (2023). Direct Preference Optimization: Your Language Model is Secretly a Reward Model. \textit{NeurIPS 2023}.

[9] Shao, Z. et al.\ (2024). DeepSeek-R1: Incentivizing Reasoning Capability in LLMs via Reinforcement Learning. \textit{arXiv:2401.02954}.

[10] Chase, H. (2022). LangChain. \url{https://github.com/langchain-ai/langchain}

}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\appendix

\section{System prompt}
\label{app:prompt}

The system prompt used for the distiller instructs the model on its role (magma equation specialist), input format (\texttt{<equation1>}/\texttt{<equation2>} tags), required output structure (\texttt{<think>}/\texttt{<answer>} blocks), and the \texttt{VERDICT:~TRUE|FALSE} format with mandatory \texttt{REASONING:} and \texttt{PROOF:} or \texttt{COUNTEREXAMPLE:} sections.

\section{Reward function details}
\label{app:reward}

The full reward function (Eq.~\ref{eq:reward}) decomposes as follows: 70\% verdict correctness provides the primary learning signal, while 5\% for \texttt{<think>} presence, 10\% for \texttt{<answer>} presence, and 10\% for structured sections (\texttt{REASONING:} + \texttt{PROOF:}/\texttt{COUNTEREXAMPLE:}) shape the output format. An unparseable response receives $R{=}0.05$ to avoid a completely flat gradient.

\section{Repository structure}
\label{app:repo}

\begin{verbatim}
Algebraic-Reasoning-Distiller/
  api/app.py                                 # FastAPI service
  container/                                 # Docker + requirements
  sair/
    agents/                                  # Specialised agents
      distiller.py                           # LLM distiller + cheat-sheet synthesis
      evaluator.py                           # Judge integration
      prover.py                              # Symbolic prover
      counterexample.py                      # Magma search
      retriever.py                           # RAG retriever
    data/
      competition/                           # Raw .txt problem dumps
      problems/                              # Parsed JSONL problem sets
      training/                              # SFT / GRPO / eval splits
      build_training_dataset.py              # Pipeline -> SFT+GRPO JSONL
      load_problems.py                       # Deduplicating JSONL loader
    scripts/
      parse_competition_problems.py          # Arrow-normalising .txt parser
      generate_cheat_sheet.py                # Balanced clustering + synthesis
    config.py                                # Hyperparameters
    equations.py                             # Parser + evaluator
    magma.py                                 # Finite magma enumeration
    graph.py                                 # Pipeline orchestration
    train_sft_distiller.py                   # SFT training
    train_grpo_distiller.py                  # GRPO training
\end{verbatim}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


\end{document}
